### 2021-01-03 paper/localpsf.tex
deleted old text from paper

% REQUIRED
\begin{abstract}
  We present an efficient matrix-free method for approximating locally translation invariant operators that have locally supported non-negative integral kernels. The idea of the method is to compute impulse responses of the operator at a collection of scattered points, then interpolate these computed impulse responses to approximate impulse responses at other arbitrary points. This results in a product-convolution approximation, which we convert to hierarchical matrix format. Once the approximation is represented in hierarchical matrix format, we perform matrix inversion and other matrix operations using fast hierarchical matrix arithmetic. Impulse responses are computed by applying the operator to a small number of Dirac comb ``batches'' of point sources. A key innovation of our method is a matrix-free procedure for choosing as many points as possible per batch, while ensuring that the supports of the impulse responses within each batch do not overlap. We apply the method to approximate Hessians in large-scale PDE-constrained inverse problem with highly informative data. Numerical results demonstrate that our method substantially outperforms existing state-of-the-art Hessian approximation methods which are based on low-rank approximation. Our method is able to form high quality approximations of high rank Hessians using only a small number of Hessian matrix-vector products.
%  Within each Dirac comb, the point sources must be spaced sufficiently far apart so that the impulse responses to those point sources do not overlap. We estimate the required spacing of point sources by applying the operator to a small number of constant, linear, and quadratic functions, then postprocessing the results.
\end{abstract}

@@

Let $\Omega \subset \mathbb{R}^d$ be a bounded domain in dimension $d=1$, $2$, or $3$, and let $\mathcal{A}:L^2(\Omega)\rightarrow L^2(\Omega)'$ be an integral operator of the form
\begin{equation}
\label{eq:kernel_representation}
(\mathcal{A}u)(v) := \int_\Omega \int_\Omega v(y) A(y,x) u(x) dx dy,
\end{equation}
with integral kernel $A:\Omega \times \Omega \rightarrow \mathbb{R}$. We focus on operators of this form which have the following properties:
\begin{description}
\item[Matrix-free:] One cannot easily access kernel entries $A(y,x)$. Access to $\mathcal{A}$ is available only through application of $\mathcal{A}$ and $\mathcal{A}^T$ to arbitrary vectors. That is, evaluation of the functions
\begin{equation*}
	u \mapsto\mathcal{A}u \quad \text{and} \quad v \mapsto\mathcal{A}^Tv.
\end{equation*} 
\item[Non-negative kernel:] For all $(y,x) \in \Omega \times \Omega$, we have
\begin{equation*}
	A(y,x) \ge 0.
\end{equation*}
\item[Local support:] $A(y, x)$ is supported near $y=x$.
\item[Local translation invariance:] If $h$ is not too large, then
\begin{equation}
	\label{eq:local_translation_invariance}
	A(y+h, x+h) \approx A(y,x).
\end{equation}
\end{description}
Local translation invariance is illustrated in Figure FIG. 
%We qualify the degree of local translation invariance with the words ``perfectly'', ``highly'', and ``poorly''.
%\begin{itemize}
%\item If equality holds in \eqref{eq:local_translation_invariance} for all $x$, $y$, and $h$, then we say $\mathcal{A}$ is \emph{perfectly translation invariant}. 
%\item If the discrepancy between left hand side and right hand side of \eqref{eq:local_translation_invariance} is small for a large set of points $x$, $y$, and $h$, then we say that $\mathcal{A}$ is \emph{highly locally translation invariant}. 
%\item If the discrepancy between left hand side and right hand side of \eqref{eq:local_translation_invariance} is large, or if the discrepancy is only small for a small set of $x$, $y$, and $h$, then we say that $\mathcal{A}$ is \emph{poorly locally translation invariant}.
%\end{itemize} 
If $\mathcal{A}$ were perfectly translation invariant (i.e., if equality held in \eqref{eq:local_translation_invariance} for all $x$, $y$, $h$) then it is straightforward to show that $\mathcal{A}$ would be a convolution operator, and the convolution kernel, $\varphi_x$, would be given by
\begin{equation}
\label{eq:convolution_kernel}
	\varphi_x(y) := A(y+x,x).
\end{equation}
The function $\varphi_x$ is the result of applying $\mathcal{A}$ to a point source, $\delta_x$, at some fixed point $x \in \Omega$, then translating the resulting impulse response function to re-center it at zero instead of $x$. Locally translation invariant operators act like convolution operators locally, but the convolution kernel, $\varphi_x$, varies as $x$ changes. We therefore approximate $\mathcal{A}$ by a spatially varying weighted sum of convolution operators of the form
\begin{equation}
\label{eq:product_convolution}
	\left(\mathcal{A}u\right)(v) \approx \left(\widetilde{\mathcal{A}}u\right)(v) := \left( v,~ \sum_{i=1}^r \varphi_i \ast \left(w_i \cdot u\right) \right)_{L^2(\Omega)},
\end{equation}
where $f \cdot g$ denotes pointwise multiplication, $f \ast g$ denotes convolution, and $\left(f, g\right)_{L^2(\Omega)}$ denotes the $L^2$ inner product on $\Omega$, for functions $f,g \in L^2(\Omega)$. Here
\begin{itemize}
	\item The local convolution kernels, $\varphi_i := \varphi_{x_i}$, are given by formula \eqref{eq:convolution_kernel}, but with $x$ replaced by $x_i$.
	\item $\{x_i\}_{i=1}^r \subset \Omega$ is a collection of points scattered throughout the domain.
	\item The functions $w_i$ are spatially varying weighting functions that are used to interpolate the convolution kernels $\varphi_i$.
\end{itemize}
Approximations of the form \eqref{eq:product_convolution} are known as \emph{product-convolution} approximations, because the action of each term in the sum consists of a pointwise product, followed by a convolution. The more locally translation invariant an operator is (i.e., the smaller the discrepancy between the left hand side and right hand side in \eqref{eq:local_translation_invariance}), the smaller the number of terms that are required in \eqref{eq:product_convolution} to achieve an accurate product-convolution approximation.

@@

%A small number of terms in \eqref{eq:product_convolution} are required to accurately approximate a highly locally translation invariant operator, while a large number of terms are required to accurately approximate a poorly locally translation invariant operator.

@@

Our motivation for this work is approximation of Hessians in distributed parameter inverse problems governed by partial differential equations (PDEs). That is, inverse problems in which one seeks to reconstruct an unknown parameter field, $m$, from noisy observations, 
\begin{equation*}
y=f(m,u(m)) + \text{noise},
\end{equation*}
which depend on  a state variable $u$. In turn, $u$ depends on $m$ implicitly through the solution of a PDE, 
\begin{equation}
\label{eq:state_pde}
0=g(m,u).
\end{equation}
Here $u(m)$ denotes the solution of the PDE \eqref{eq:state_pde} as a function of $m$.
%
%That is, inverse problems in which one seeks to reconstruct an unknown parameter field, $m$, from noisy measurements,
%\begin{equation*}
%	y = B(m,u) + \xi,
%\end{equation*}
%of a state variable, $u$, which depends on $m$ indirectly through the solution of a PDE,
%\begin{equation}
%\label{eq:state_equation}
%	0=F(m,u).	
%\end{equation}
%Here $\xi$ is the noise. The observations $y$ are known, while $m$, $u$, and $\xi$ are unknown. In the deterministic setting, the inverse problem of finding $m$ is commonly framed as a least squares minimization problem
%\begin{equation}
%\label{eq:minimization_problem}
%	\min_m \quad J_d(m) + R(m),
%\end{equation}
%where
%\begin{equation*}
%	J_d(m) := \frac{1}{2}\|y - B(m,u(m))\|_{\Gamma}^2
%\end{equation*}
%is a data misfit term, $\Gamma$ is the noise covariance, $R(m)$ is a regularization term, and $u(m)$ denotes the solution $u$ of the PDE \eqref{eq:state_equation} as a function of $m$, . 
%
%Approximation of the Hessian,
%\begin{equation*}
%	\mathcal{H} = \mathcal{H}_d + \mathcal{H}_R := \frac{d^2 J_d}{dm^2} + \frac{d^2 R}{dm^2},
%\end{equation*}
In the deterministic approach to inverse problems, one typically finds $m$ as the solution to the minimization problem 
\begin{equation}
\label{eq:minimization_problem}
\min_m \quad \frac{1}{2}\|y - f(m,u(m))\|_W^2 + R(m),
\end{equation}
where $\|\cdot\|_W$ is weighted norm which depends on the noise covariance, and $R(m)$ is a regularization term. Approximation of the Hessian of the objective function, $\mathcal{H}$, allows for fast solution of \eqref{eq:minimization_problem} via Newton-type methods. Hessian approximations are also central to many methods for uncertainty quantification in Bayesian statistical approaches to the inverse problem, because $\mathcal{H}^{-1}$ locally approximates the Bayesian posterior covariance for $m$. 
We only have access to $\mathcal{H}$ via its action on an arbitrary vectors $v$, i.e., evaluation of the map $v \mapsto \mathcal{H} v$. Evaluating $v \mapsto \mathcal{H} v$ is an expensive process that involves solving two auxiliary linear PDEs.

%The regularization term in the Hessian, $\mathcal{H}_R := \frac{d^2 R}{dm^2}$, is typically an elliptic differential operator, and is therefore easy to work with. The data misfit term, $\mathcal{H}_d = \frac{d^2 J}{dm^2}$, is difficult to work with. The data misfit term $\mathcal{H}_d$ is not directly accessible because it depends on $u$, which depends on $m$ implicitly through the solution of the PDE \eqref{eq:state_equation}. 
%
%We only have access to $\mathcal{H}_d$ via its action on an arbitrary vectors $v$, i.e., evaluation of the map $v \mapsto \mathcal{H} v$. Evaluating $v \mapsto \mathcal{H} v$ is an expensive process that involves solving two auxiliary PDEs. We show how to evaluate $v \mapsto \mathcal{H} v$ in Appendix ADSF.

The most popular existing Hessian approximation methods are based on forming a low rank approximation of the data misfit term in the Hessian, or the data misfit term preconditioned by the prior term [CITE LOTS OF STUFF]. Either the Lanczos method or the randomized singular value decomposition [CITE RSVD] are used to perform the low rank approximation using only matrix-vector products. These methods suffer from a ``data predicament''---as more informative data is used in the inversion, the numerical rank of the data misfit term in the Hessian grows, so more matrix-vector products are required to form the aforementioned low rank approximation. The ideal scenario from a scientific perspective (highly informative data) is the worst case scenario from a computational perspective (large computational cost) [CITE MY DISSERTATION]. 
%In our numerical results, we show that our operator approximation method can efficiently approximate the data misfit term in the Hessian in certain PDE constrained inverse problems. 
The method in this paper offers a \emph{data-scalable} alternative to conventional low-rank Hessian approximation methods, because high-rank approximations of an operator can be formed using a small number of matrix-vector products.

@@

We build upon the class of product convolution approximations in which the functions, $\varphi_i$, are impulse responses of $\mathcal{A}$ to point sources at a collection of points $x_i$ [CITE BIG LIST]. 
%The primary question in methods of this type are how to choose the points $x_i$, and how to choose the weighting functions $w_i$.
A popular choice for methods in this class is to choose the points $x_i$ to be nodes in a regular grid, and interpolate the functions $\varphi_i$ with piecewise linear interpolation [CITE NAGY], or splines [CITE SPLINES]. With a regular grid, a large number of points $x_i$ is typically required to achieve an accurate product-convolution approximation. This is computationally prohibitive in our applications, because computing the functions $\varphi_i$ requires one matrix vector product with $\mathcal{A}$ for each point $x_i$, and these matrix vector products are computationally costly. In our previous work, we reduced the computational cost by starting with a coarse grid of points $x_i$, then adaptively refining the grid in the regions where the error in the approximation is large [CITE, CITE MY PAPER]. But even with adaptive refinement, many matrix vector products with $\mathcal{A}$ may be required. In this paper, rather than using adaptive refinement, we instead reduce the computational cost by picking points $x_i$ such that many $\varphi_i$ are computed with each matrix vector product. We do this by taking advantage of the fact that the functions $\varphi_i$ are locally supported; existing methods typically do not require this local support property.

@@

%The goal of these existing papers is to pick points $x_i$ so that the required the number of $\varphi_i$ is minimized. Our goal here is to minimize the required number of matrix vector products with $\mathcal{H}$.

%We do this by taking advantage of the local support property of $\mathcal{A}$.

%SVIR low rank: sample columns of matrix

%Pros and cons of different point picking strategies depend on the application of interest.

%In astronomy image deblurring applications, the point source positions typically correspond to locations of stars in an image, which are fixed.

%In [CITE], Escande and Weiss explored the idea of approximating linear operators from scattered impulse responses using a product-convolution scheme of the same form as \eqref{eq:product_convolution}.

%In our previous work [CITE MY PAPER], we choose the point locations with an adaptive grid. However,

%Nagy's paper: breaking domain into regular grid, using piecewise linear interpolation.

@@

%H-matrices for Hessians: Ilonas paper with KAUST people

%Demanet matrix probing can be interpreted as convolution-product.

%Resolution analysis: uses matvecs with random functions to estimate spatial covariance. Was inspiration for us.

%Getting many impulse responses at once (batching).

%Geometry-aware weighting functions vs radial basis functions. Kriging.

%\subsection{Operator access levels}
%
%
%%In order to solve inverse or optimization problems governed by PDEs, we need to solve linear systems with the Hessian as the coefficient matrix [cite cite]. In order to do this efficiently, we have access to only Hessian-vector products. That is, we have the ability to compute the functions $x \mapsto \mathcal{H} x$ and $x \mapsto \mathcal{H}^T x$ for arbitrary vectors $x$. Therefore 
%
%A linear operator may be classified by the way in which we may access the operator. 
%
%
%\begin{itemize}
%	\item \emph{Level 1:} A matrix representation of the operator, $\mathbf{A}$, is stored in memory in dense or sparse format.
%	\item \emph{Level 2:} We can efficiently compute arbitrary matrix entries $\mathbf{A}_{ij}$ at will, but the matrix $\mathbf{A}$ is not stored in memory.
%	\item \emph{Level 3:} Access to $\mathbf{A}$ is only available through an algorithm that computes matrix-vector products $\mathbf{u} \mapsto \mathbf{A}\mathbf{u}$ and $\mathbf{u} \mapsto \mathbf{A}^T \mathbf{u}$.
%\end{itemize}
%In large-scale scientific computing, 
%
%
%As discussed in Section \ref{sec:PDE_hessian_motivation}, in inverse problems governed by PDEs, matrix entries of the Hessian are not easily computable since getting one entry would require computing an entire column of the Hessian, then throwing away all but one entry in the column. Computing a single column costs at least two PDE solves. Many such entries are needed to approximate the Hessian in inverse problems with highly informative data (i.e., for Hessians with large numerical rank) [cite cite].
%
% of the implicit definition of $u(m)$. 
%
%Hierarchical matrices (H, H2, HSS, HBS, HODLR), quantized tensor trains, hierarchical interpolative factorization, STRUMPACK (LBNL), 
%
%Level 1: not relevant in large-scale problems
%State clearly: we do not use Level 1 at all.
%This paper: Level 3 -> Level 2
%H-matrix: solve on level 2
%
%%\subsection{Existing work}
%%
%%
%%\begin{itemize}
%%	\item CITE ESCANDE, others, for a reviews of the subject.
%%	\item Resolution analysis seismic papers: try to figure out size of $\varphi_x$ by matvec with random function, post-processing to get autocorrelation length. They did the constant matvec volume thing. Their goal is to get a sense of local length scales that can be resolved in the inverse problem. They do not build preconditioner
%%	\item Escande Weiss scattered response paper. They interpolate scattered impulse responses using radial basis functions and study how the geometry of the points affects the approximation accuracy. Matrix-free / getting many points at once is not a concern of them. The applications they consider one has direct access to the impulse response, e.g., through formula for the kernel. 
%%	\item Imaging problems: ``matvec'' requires performing an experiment
%%	\item pseudodifferential operator equivalence
%%	\item Laundry list of applications
%%	\item scattered impulse response of escande weiss paper: radial basis functions for approximation in $y$. How to determine if the impulse responses are far enough away from each other?
%%	\item my convolution paper: compute impulse responses one at a time, one impulse response per matvec: too slow! Here we compute many impulse responses using only 1 matvec.
%%	\item uniformity of smoothness condition
%%\end{itemize}

\section{Constructing the approximation}
\label{sec:method}
This section describes how we compute the convolution kernels, $\varphi_i$, and weighting functions, $w_i$, for the product-convolution approximation, $\widetilde{\mathcal{A}} \approx \mathcal{A}$, given in \eqref{eq:product_convolution}.
\begin{itemize}
\item We extract $\varphi_{x_i}$ for many points $x_i$ by applying $\mathcal{A}$ to Dirac combs associated with ``batches'' of points (Section \ref{sec:get_impulse_response}). 
\item We use a greedy algorithm to choose as many points as possible per batch, while ensuring that the points in each batch are not too close to each other, and are not too close to $\partial \Omega$ (Section \ref{sec:greedy_point_selection}). 
\item Ensuring that the points $x_i$ are well separated from each other requires a-priori estimates of the supports of the functions $\langle\mathcal{A},\delta_{x}\rangle^*$ (Section \ref{eq:mean_and_covariance_estimation}). 
\item The weighting functions are defined as the smoothest functions (in a least-squares sense) that interpolate the points $x_i$; constructing the weighting functions requires solving two Poisson PDEs per point $x_i$ (Section \ref{sec:weighting_functions}).
\end{itemize}
The complete algorithm for constructing $\widetilde{\mathcal{A}}$ is shown in Algorithm \ref{alg:construct_Atilde}.

\begin{algorithm2e}
	\SetAlgoNoLine
	\SetKwInOut{Input}{Input}
	\SetKwInOut{Output}{Output}
	%	\Input{Function $v \mapsto \mathcal{H}^Tv$}

	%	\mbox{}\\

	%	\Output{Spatially varying mean $\mu:\Omega \rightarrow \mathbb{R}^d$, \\ 
	%		Spatially varying covariance $\Sigma:\Omega \rightarrow \mathbb{R}^{d \times d}$}
	%	\Begin{
	{	
		Compute spatially varying mean, $\mu$, and covariance, $\Sigma$, using Algorithm \ref{alg:varhpi_mean_cov}.

@@

		%		\Return{$\mu$, $\Sigma$}
	}
	\caption{Construct product-convolution approximation $\widetilde{\mathcal{A}}$}
	\label{alg:construct_Atilde}
\end{algorithm2e}

%The function $\varphi_x(y)$ is undefined when $x + y \notin \Omega$, which can occur even if $x \in \Omega$ and $y \in \Omega$. Extending $\varphi_x$ by zero makes it defined for all $y \in \mathbb{R}^d$, but this extension is only reasonable if $x$ is far enough from $\partial \Omega$ that $\varphi_x(y)$ decays to zero, as a function of $y$, before $y$ reaches $\partial \Omega$. We therefore choose exclude sample points $x_i$ if the ellipsoid associated with that point overlaps the boundary.

@@

%The support of $\langle\mathcal{A},\delta_{x}\rangle^*$ is estimated to be contained within a small number of standard deviations of the mean for a Gaussian distribution with the same mean and covariance as the normalized version of $\langle\mathcal{A},\delta_{x}\rangle^*$.

%Let 
%\begin{equation*}
%	\rho_x := \langle\mathcal{A}, \delta_x\rangle^* \big/ \alpha(x)
%\end{equation*}
%denote the normalized version of $\langle\mathcal{A}, \delta_x\rangle^*$, where $\alpha:\Omega \rightarrow \mathbb{R}$ is the volume function
%\begin{equation*}
%	\alpha(x) := \int_{\Omega} \langle\mathcal{A}, \delta_x\rangle^*(z) dz.
%\end{equation*}
%Note that $\rho_x$ is a probability distribution on $\Omega$.
%
%In Proposition \ref{thm:vol_mean_cov}, we see that $\alpha(x)$ may be computed for all $x$ simultaneously by applying $\mathcal{A}^T$ to a constant function. We also see that the spatially varying mean, $\mu(x)$, and covariance, $\Sigma(x)$, of $\rho_x$, defined as
%\begin{align*}
%	\mu(x) :=& \int_\Omega z \rho_x(z) dz \\
%	\Sigma(x) :=& \int_\Omega (z - \mu(x))(z - \mu(x))^T \rho_x(z) dz
%\end{align*}
%may be computed for all $x$ simultaneously by applying $\mathcal{A}^T$ to $d$ linear functions, and $d(d+1)/2$ quadratic functions, respectively. In Algorithm \ref{alg:varhpi_mean_cov} we describe an algorithm for computing the volume function, mean, and covariance based on the results of Proposition \ref{thm:vol_mean_cov}.

@@

\begin{algorithm2e}
	\SetAlgoNoLine
	\SetKwInOut{Input}{Input}
	\SetKwInOut{Output}{Output}
%	\Input{Function $v \mapsto \mathcal{H}^Tv$}

%	\mbox{}\\

%	\Output{Spatially varying mean $\mu:\Omega \rightarrow \mathbb{R}^d$, \\ 
%		Spatially varying covariance $\Sigma:\Omega \rightarrow \mathbb{R}^{d \times d}$}
%	\Begin{
	{	
		Form constant function $C(x)=1$

@@

%		\Return{$\mu$, $\Sigma$}
	}
	\caption{Compute volume $\alpha$, mean $\mu$, and covariance $\Sigma$}
	\label{alg:varhpi_mean_cov}
\end{algorithm2e}

@@

% sample points $x_i$ and $x_j$ should be far enough apart from each other that the supports of $\left\langle\mathcal{A}, \delta_{x_i}\right\rangle^*$ and $\left\langle\mathcal{A}, \delta_{x_j}\right\rangle^*$ do not overlap.

%Since it is too expensive compute the support of $\left\langle\mathcal{A}, \delta_{x}\right\rangle^*$ directly, we instead estimate that the support of $\left\langle\mathcal{A}, \delta_{x}\right\rangle^*$ is contained in the ellipsoid
%\begin{equation}
%\label{eq:support_ellipsoid}
%	E_x := \{x' \in \Omega: (x' - \mu(x))^T \Sigma(x)^{-1} (x' - \mu(x)) \le \tau^2\},
%\end{equation}
%where $\tau$ is a small number of standard deviations; we use $\tau=3$. The ellipsoid $E_x$ is the set of points within $\tau$ standard deviations of the mean for the Gaussian distribution that has the same mean and covariance as $\left\langle\mathcal{A}, \delta_{x}\right\rangle^*$. The fraction of the mass of $\left\langle\mathcal{A}, \delta_{x}\right\rangle^*$ residing outside of $E_x$ is less than $1/\tau^2$ by Chebyshev's inequality, though we find this bound is conservative and far less mass resides in this region in practice.

%We call $E_x$ the \emph{unit ellipsoid} with mean $\mu(x)$ and covariance $\Sigma(x)/\tau^2$.

Starting from a finite set of candidate points $P$, we select batches of sample points, $S_b$, using a greedy algorithm. To build a batch $S_b$, first we initialize $S_b$ as an empty set. Then we select a candidate point $p \in P$ that is the farthest away from all points in previous sample point batches. That is, $p$ is the maximizer of the following optimization problem:
\begin{equation*}
\max_{p \in P} \min_{x \in S_1 \cup \dots \cup S_{b-1}} \|p - x\|.
\end{equation*}
Once $p$ is selected, we remove $p$ from $P$. If $p$ is sufficiently far from all of the previously chosen points in the current batch, in the sense that $E_p \cap E_q = \{\}$ for all $q \in S_b$, and if $p$ is sufficiently far from the boundary, in the sense that $E_p \cap \partial \Omega = \{\}$, then we add $p$ to $S_b$. Otherwise we discard $p$. This process repeats until there are no more points in $P$.  This is detailed in Algorithm \ref{alg:point_choice}. We determine whether $E_p \cap E_q = \{\}$ using the fast ellipsoid intersection test described in Appendix \ref{sec:fast_ellipsoid_intersection_test}.

%We select a batches of sample points, $S_b$, from a finite set of candidate points $P$ using a greedy algorithm. Starting from an empty set, we build $S_b$ from candidate points $p$ from $P$, one point at a time. The current candidate point, $p$, is the point in $P$

@@

%		\If{$\text{point\_is\_acceptable}(p, S_b)$}{
		\If{$E_p \cap E_q \neq \{\}$ for all $q \in S_b$ and $E_p \cap \partial \Omega = \{\}$}{

@@

%Note that the cluster trees and block cluster tree are computed using only the geometric information about the degree of freedom locations, and do not depend on the values of the matrix entries of $\widetilde{\mathcal{A}}$.

@@

\section{Interpolating impulse response samples old}

>>>>

% REQUIRED
\begin{abstract}
  We present an efficient matrix-free method for approximating locally translation invariant operators that have locally supported non-negative integral kernels. The idea of the method is to compute impulse responses of the operator at a collection of scattered points, then interpolate these computed impulse responses to approximate impulse responses at other arbitrary points. This results in a product-convolution approximation, which we convert to hierarchical matrix format. Once the approximation is represented in hierarchical matrix format, we perform matrix inversion and other matrix operations using fast hierarchical matrix arithmetic. Impulse responses are computed by applying the operator to a small number of Dirac comb ``batches'' of point sources. A key innovation of our method is a matrix-free procedure for choosing as many points as possible per batch, while ensuring that the supports of the impulse responses within each batch do not overlap. We apply the method to approximate Hessians in large-scale PDE-constrained inverse problem with highly informative data. Numerical results demonstrate that our method substantially outperforms existing state-of-the-art Hessian approximation methods which are based on low-rank approximation. Our method is able to form high quality approximations of high rank Hessians using only a small number of Hessian matrix-vector products.
\end{abstract}

@@

Let $\Omega \subset \mathbb{R}^d$ be a bounded domain in dimension $d=1$, $2$, or $3$, and let $\mathcal{A}:L^2(\Omega)\rightarrow L^2(\Omega)'$ be an integral operator of the form
\begin{equation}
\label{eq:kernel_representation}
(\mathcal{A}u)(v) := \int_\Omega \int_\Omega v(y) A(y,x) u(x) dx dy,
\end{equation}
with integral kernel $A:\Omega \times \Omega \rightarrow \mathbb{R}$. We focus on operators of this form which have the following properties:
\begin{description}
\item[Matrix-free:] One cannot easily access kernel entries $A(y,x)$. Access to $\mathcal{A}$ is available only through application of $\mathcal{A}$ and $\mathcal{A}^T$ to arbitrary vectors. That is, evaluation of the functions
\begin{equation*}
	u \mapsto\mathcal{A}u \quad \text{and} \quad v \mapsto\mathcal{A}^Tv.
\end{equation*} 
\item[Non-negative kernel:] For all $(y,x) \in \Omega \times \Omega$, we have
\begin{equation*}
	A(y,x) \ge 0.
\end{equation*}
\item[Local support:] $A(y, x)$ is supported near $y=x$.
\item[Local translation invariance:] If $h$ is not too large, then
\begin{equation}
	\label{eq:local_translation_invariance}
	A(y+h, x+h) \approx A(y,x).
\end{equation}
\end{description}
Local translation invariance is illustrated in Figure FIG. If $\mathcal{A}$ were perfectly translation invariant (i.e., if equality held in \eqref{eq:local_translation_invariance} for all $x$, $y$, $h$) then it is straightforward to show that $\mathcal{A}$ would be a convolution operator, and the convolution kernel, $\varphi_x$, would be given by
\begin{equation}
\label{eq:convolution_kernel}
	\varphi_x(y) := A(y+x,x).
\end{equation}
The function $\varphi_x$ is the result of applying $\mathcal{A}$ to a point source, $\delta_x$, at some fixed point $x \in \Omega$, then translating the resulting impulse response function to re-center it at zero instead of $x$. Locally translation invariant operators act like convolution operators locally, but the convolution kernel, $\varphi_x$, varies as $x$ changes. We therefore approximate $\mathcal{A}$ by a spatially varying weighted sum of convolution operators of the form
\begin{equation}
\label{eq:product_convolution}
	\left(\mathcal{A}u\right)(v) \approx \left(\widetilde{\mathcal{A}}u\right)(v) := \left( v,~ \sum_{i=1}^r \varphi_i \ast \left(w_i \cdot u\right) \right)_{L^2(\Omega)},
\end{equation}
where $f \cdot g$ denotes pointwise multiplication, $f \ast g$ denotes convolution, and $\left(f, g\right)_{L^2(\Omega)}$ denotes the $L^2$ inner product on $\Omega$, for functions $f,g \in L^2(\Omega)$. Here
\begin{itemize}
	\item The local convolution kernels, $\varphi_i := \varphi_{x_i}$, are given by formula \eqref{eq:convolution_kernel}, but with $x$ replaced by $x_i$.
	\item $\{x_i\}_{i=1}^r \subset \Omega$ is a collection of points scattered throughout the domain.
	\item The functions $w_i$ are spatially varying weighting functions that are used to interpolate the convolution kernels $\varphi_i$.
\end{itemize}
Approximations of the form \eqref{eq:product_convolution} are known as \emph{product-convolution} approximations, because the action of each term in the sum consists of a pointwise product, followed by a convolution. The more locally translation invariant an operator is (i.e., the smaller the discrepancy between the left hand side and right hand side in \eqref{eq:local_translation_invariance}), the smaller the number of terms that are required in \eqref{eq:product_convolution} to achieve an accurate product-convolution approximation.

@@

@@

Our motivation for this work is approximation of Hessians in distributed parameter inverse problems governed by partial differential equations (PDEs). That is, inverse problems in which one seeks to reconstruct an unknown parameter field, $m$, from noisy observations, 
\begin{equation*}
y=f(m,u(m)) + \text{noise},
\end{equation*}
which depend on  a state variable $u$. In turn, $u$ depends on $m$ implicitly through the solution of a PDE, 
\begin{equation}
\label{eq:state_pde}
0=g(m,u).
\end{equation}
Here $u(m)$ denotes the solution of the PDE \eqref{eq:state_pde} as a function of $m$. In the deterministic approach to inverse problems, one typically finds $m$ as the solution to the minimization problem 
\begin{equation}
\label{eq:minimization_problem}
\min_m \quad \frac{1}{2}\|y - f(m,u(m))\|_W^2 + R(m),
\end{equation}
where $\|\cdot\|_W$ is weighted norm which depends on the noise covariance, and $R(m)$ is a regularization term. Approximation of the Hessian of the objective function, $\mathcal{H}$, allows for fast solution of \eqref{eq:minimization_problem} via Newton-type methods. Hessian approximations are also central to many methods for uncertainty quantification in Bayesian statistical approaches to the inverse problem, because $\mathcal{H}^{-1}$ locally approximates the Bayesian posterior covariance for $m$. 
We only have access to $\mathcal{H}$ via its action on an arbitrary vectors $v$, i.e., evaluation of the map $v \mapsto \mathcal{H} v$. Evaluating $v \mapsto \mathcal{H} v$ is an expensive process that involves solving two auxiliary linear PDEs.

The most popular existing Hessian approximation methods are based on forming a low rank approximation of the data misfit term in the Hessian, or the data misfit term preconditioned by the prior term [CITE LOTS OF STUFF]. Either the Lanczos method or the randomized singular value decomposition [CITE RSVD] are used to perform the low rank approximation using only matrix-vector products. These methods suffer from a ``data predicament''---as more informative data is used in the inversion, the numerical rank of the data misfit term in the Hessian grows, so more matrix-vector products are required to form the aforementioned low rank approximation. The ideal scenario from a scientific perspective (highly informative data) is the worst case scenario from a computational perspective (large computational cost) [CITE MY DISSERTATION]. The method in this paper offers a \emph{data-scalable} alternative to conventional low-rank Hessian approximation methods, because high-rank approximations of an operator can be formed using a small number of matrix-vector products.

@@

We build upon the class of product convolution approximations in which the functions, $\varphi_i$, are impulse responses of $\mathcal{A}$ to point sources at a collection of points $x_i$ [CITE BIG LIST]. A popular choice for methods in this class is to choose the points $x_i$ to be nodes in a regular grid, and interpolate the functions $\varphi_i$ with piecewise linear interpolation [CITE NAGY], or splines [CITE SPLINES]. With a regular grid, a large number of points $x_i$ is typically required to achieve an accurate product-convolution approximation. This is computationally prohibitive in our applications, because computing the functions $\varphi_i$ requires one matrix vector product with $\mathcal{A}$ for each point $x_i$, and these matrix vector products are computationally costly. In our previous work, we reduced the computational cost by starting with a coarse grid of points $x_i$, then adaptively refining the grid in the regions where the error in the approximation is large [CITE, CITE MY PAPER]. But even with adaptive refinement, many matrix vector products with $\mathcal{A}$ may be required. In this paper, rather than using adaptive refinement, we instead reduce the computational cost by picking points $x_i$ such that many $\varphi_i$ are computed with each matrix vector product. We do this by taking advantage of the fact that the functions $\varphi_i$ are locally supported; existing methods typically do not require this local support property.

@@

@@

\section{Constructing the approximation}
\label{sec:method}
This section describes how we compute the convolution kernels, $\varphi_i$, and weighting functions, $w_i$, for the product-convolution approximation, $\widetilde{\mathcal{A}} \approx \mathcal{A}$, given in \eqref{eq:product_convolution}.
\begin{itemize}
\item We extract $\varphi_{x_i}$ for many points $x_i$ by applying $\mathcal{A}$ to Dirac combs associated with ``batches'' of points (Section \ref{sec:get_impulse_response}). To reduce memory usage, all kernels $\varphi_i$ associated with a given batch, $b$, are stored in a single function, $\eta^b$, rather than as individual functions.
\item We use a greedy algorithm to choose as many points as possible per batch, while ensuring that the points in each batch are not too close to each other, and are not too close to $\partial \Omega$ (Section \ref{sec:greedy_point_selection}). 
\item We ensure that the points $x_i$ are well separated from each other by forming a-priori estimates of the supports of the functions $\langle\mathcal{A},\delta_{x}\rangle^*$ (Section \ref{eq:mean_and_covariance_estimation}). 
\item For the weighting functions, we use the smoothest functions (in a least-squares sense) that interpolate the points $x_i$; constructing the weighting functions requires solving two Poisson PDEs per point $x_i$ (Section \ref{sec:weighting_functions}).
\end{itemize}
The complete algorithm for constructing $\widetilde{\mathcal{A}}$ is shown in Algorithm \ref{alg:construct_Atilde}.

\begin{algorithm2e}
	\SetAlgoNoLine
	\SetKwInOut{Input}{Input}
	\SetKwInOut{Output}{Output}
	{	
		Compute spatially varying mean, $\mu$, and covariance, $\Sigma$, using Algorithm \ref{alg:varhpi_mean_cov}.

@@

	}
	\caption{Construct product-convolution approximation $\widetilde{\mathcal{A}}$}
	\label{alg:construct_Atilde}
\end{algorithm2e}

@@

@@

\begin{algorithm2e}
	\SetAlgoNoLine
	\SetKwInOut{Input}{Input}
	\SetKwInOut{Output}{Output}
	{	
		Form constant function $C(x)=1$

@@

	}
	\caption{Compute volume $\alpha$, mean $\mu$, and covariance $\Sigma$}
	\label{alg:varhpi_mean_cov}
\end{algorithm2e}

@@

Starting from a finite set of candidate points $P$, we select batches of sample points, $S_b$, using a greedy algorithm. To build a batch $S_b$, first we initialize $S_b$ as an empty set. Then we select a candidate point $p \in P$ that is the farthest away from all points in previous sample point batches. That is, $p$ is a maximizer of the following optimization problem:
\begin{equation*}
\max_{p \in P} \min_{x \in S_1 \cup \dots \cup S_{b-1}} \|p - x\|.
\end{equation*}
Once $p$ is selected, we remove $p$ from $P$. If $p$ is sufficiently far from all of the previously chosen points in the current batch, in the sense that $E_p \cap E_q = \{\}$ for all $q \in S_b$, and if $p$ is sufficiently far from the boundary, in the sense that $E_p \cap \partial \Omega = \{\}$, then we add $p$ to $S_b$. Otherwise we discard $p$. This process repeats until there are no more points in $P$.  This is detailed in Algorithm \ref{alg:point_choice}. We determine whether $E_p \cap E_q = \{\}$ using the fast ellipsoid intersection test described in Appendix \ref{sec:fast_ellipsoid_intersection_test}.

@@

		\If{$E_p \cap E_q \neq \{\}$ for all $q \in S_b$ and $E_p \cap \partial \Omega = \{\}$}{

@@

@@

\section{Numerical results}
\label{sec:numerical_results}
